---
output:
  html_document: default
  pdf_document: default
editor_options: 
  chunk_output_type: inline
---
# **A FHIR PIT tutorial** {.unnumbered}

|                                                                   |
|-------------------------------------------------------------------|
| title: "FHIR PIT: FHIR and environmental exposures integration tool" |
| subtitle: "Tutorial"                                              |
| date: "April 2, 2024"                                        |
| author: "Juan Garcia, Kara Fecho, Hong Yi"                                              |
| format: html                                                      |

*Date*: April 2, 2024

*Author(s)*: Juan Garcia, Kara Fecho, Hong Yi

# FHIR PIT: FHIR and environmental exposures integration tool: A tutorial

**Key Terms**: Data Integration, Environmental exposures, EHR, FHIR

**Programming Language(s)**: R

## Introduction

This tutorial provides example code in `R`:

1.  To setup FHIR PIT.

2.  To analyze the results from the FHIR PIT output.

### Disclaimer

The data used in this tutorial is synthetic. If you would like to incorporate your own consider a secure enclave.

Note the FHIRPIT application may be memory intensive depending on the size of the input data. For example, this demo  requires 25Gb to run successfully.

### Motivation

Environmental exposures are increasingly recognized as important to consider when conducting research in numerous scientific fields, including biomedicine, social science, and political science. Unfortunately, associating environmental exposure data with subject-level data is difficult due to varying spatiotemporal resolutions. FHIR PIT is a tool to link EHR records in FHIR format with environmental exposures. 

### Background 

FHIR PIT (HL7 Fast Healthcare Interoperability Resources Patient data Integration Tool) is an open-source data integration pipeline to support subject-level research on social and environmental determinants of health and disease.

In essence, the pipeline is a sequence of transformations configured from a YAML file.
The transformation steps join FHIR patient records with environmental exposure estimates using patient geocodes (i.e., primary residence) and date(s) of healthcare visit as the spatiotemporal links. At the end of each linkage step, FHIR PIT saves the linked tables to an intermediate folder to support rapid re-execution, should any given step fail to complete (e.g., lack of disk space). To support multiple healthcare visits, FHIR PIT vectorizes FHIR patient records by grouping daily FHIR domains (e.g., Condition, Laboratory, Medication, Procedures) and counting them. The features associated with the patient are then grouped by user-defined study periods (e.g., yearly) and aggregated with user-defined statistics (e.g., counts, mean, median, first measurement, last measurement).

### Outline

This tutorial includes the following steps:

-   [Step 0. Setup](#step-0)

-   [Step 1. Run FHIR PIT](#step-1)

-   [Step 2. Analyze results](#step-2)

## Tutorial

```{r}
# Loading library
library(ggplot2)
```


### Step 0. Setup {#step-0}

Install Docker. Follow instructions in [Docker installation](https://docs.docker.com/engine/install/) documentation for your operating system.

Assign WORK_DIR the directory you would like to save FHIR PIT and its outputs to.

```{r}
work_dir <- "/Users/jj/Desktop/FHIR-PIT"
Sys.setenv(WORK_DIR = work_dir)
```

Clone FHIR PIT into the WORK_DIR directory.

```{bash}
mkdir -p $WORK_DIR
git clone --recursive https://github.com/ExposuresProvider/FHIR-PIT.git $WORK_DIR
```

### Step 1. Execute FHIR-PIT {#step-1}

```{bash}
echo "$WORK_DIR/data/output"
```

Run FHIR PIT in a docker container with -v. This will allow the output data to persist and be easily accessible by users on the host.

```{bash}
docker run -m="25g" -v "$WORK_DIR/data/output":"$WORK_DIR/data/output" renci/fhir-pit:1.0 
```

### Step 2. Access and analyze results {#step-2}

The main outputs to consider are in ICEES, ICEES2 and ICEES2_dei directories. Each directory contains one or multiple csv files with information at different granularities. 


```{r}
year_ <- 2010
basepath <- file.path(work_dir, "data", "output")
icees2dei_url <- file.path(basepath,"icees2_dei",  sprintf("%spatient_deidentified", year_))
```

ICEES contains a CSV per subject. Each CSV consists of the subject’s visits, concatenated with the corresponding environmental exposures that occurred during that day. If a subject has multiple visits per day, the transformation PreprocPerPatSeriesToVector aggregates multiple daily visits by counting how many times a drug/diagnose appeared. The directory for each subject is indexed by the "patient_num" column. ICEES2 contains a single CSV with the aggregation of all patients  grouped by (Subject, study period). Lastly,ICEES2DEI contains the same information but deidentified. This is the recommended output for data sharing. For exposition purposes we focus on ICEES2DEI and reorder its columns.

```{r}
icees2dei_colorder <- scan("icees2dei_column_order.txt", what="", sep="\n")
icees2dei <- read.csv(icees2dei_url, header = TRUE)[, icees2dei_colorder]
```

```{r}
icees2dei
```

The output of ICEES2DEI may be used to fo further analysis. For instance, below we plot `AvgDailyPM2.5Exposure` vs `AvgDailyOzoneExposure` to identify any correlation between average daily ozone exposure and average daily PM2.5 exposure.

```{r}
ggplot(icees2dei,aes(y=AvgDailyPM2.5Exposure,x=AvgDailyOzoneExposure)) + geom_point(color='blue') +
  geom_smooth(method = "lm", se = FALSE)
```


## Considerations
FHIR PIT currently assumes a patient or subject resides in a single geolocation across a defined study period.
FHIR records with exposure data through patient’s address, and environmental data through patient’s address, study period and date-of-visit. 

## Concluding Remarks
This tutorial demonstrates how to execute FHIR PIT and analyze its output. The data used in this tutorial is synthetic for demostranstion purposes. The user may download the complete environmental exposures dataset from [link] and add their own FHIR data for realistic analysis. To add data, update the  `data/FHIR/all` directory. 

## Resources

[1] H. Xu et al., “FHIR PIT: an open software application for spatiotemporal integration of clinical data and environmental exposures data,” BMC Med Inform Decis Mak, vol. 20, no. 1, p. 53, Dec. 2020, doi: 10.1186/s12911-020-1056-9.

[2] J.J. Garcia, K. Fecho, H. Yi, and H. Xu, “FHIR-PIT: FHIR Patient data Integration Tool.” [Online]. Available: https://github.com/ExposuresProvider/FHIR-PIT

[3] Garcia, Juan, “FHIR-PIT: Transformations documentation” [Online]. Available: https://github.com/ExposuresProvider/FHIR-PIT/blob/contained_demo/spark/FHIR-PIT-Transformations-Documentation.pdf

